A flexible integer linear programming formulation for scheduling clinician on-call service in hospitals

### 2019-09-13 paper/paper.problem.tex
paragraph spacing
tables & figures cleanup

At St.\ Michael's Hospital, the division of infectious diseases (ID) offers
general ID and HIV consultation on inpatients as two parallel services. Each
service provides clinical care throughout the year, during regular work weeks as
well as weekends and holidays. The clinicians in the division typically receive
a schedule in advance, outlining their on-call service dates for the full year.
In the yearly schedule each clinician is assigned to blocks of regular work
weeks and weekends. Each block corresponds to two consecutive work weeks. Apart
from long (holiday) weekends, a work week starts on Monday at 8 A.M. and ends on
Friday at 5 P.M. Accordingly, weekend service starts on Friday at 5 P.M., and
ends on Monday at 8 A.M. During the weekend, ID and HIV consultation services
are combined and provided by one clinician. During the regular work week the ID
and HIV services are led by one clinician each. \\

Several constraints are placed on the on-call assignments. First, each clinician
has limits on the number of blocks they can and must work during the year,
depending on the type of consultation. For instance, a clinician might have to
provide 3-5 blocks of general ID consultation as well as 2-3 blocks of HIV
consultation throughout the year. These limits may change from year to year as
the number of clinicians in the department changes. Moreover, the schedule
should not assign a clinician to work for two consecutive blocks or two
consecutive weekends in a row. The schedule should also distribute both regular
and holiday weekends equally among all clinicians. \\

@@

\begin{table}[h]
	\centering
	\begin{tabular}{ l l l }
		\hline
		\textbf{Constraint Name} & \textbf{Description}                               
		&
		\textbf{Type} \\ \hline
		Block Coverage (BC)           & \makecell[l]{each service needs to have
			exactly \\ one clinician that covers any given block}                           
		& Hard          \\ \hline
		Weekend Coverage (WC)        & \makecell[l]{every weekend needs to have
			exactly \\ one clinician that covers it}                                        
		& Hard          \\ \hline
		Min/Max (MM)                  & \makecell[l]{for a given service, each
			clinician can only \\ work between the minimum and maximum \\ number of allowed
			blocks} & Hard          \\ \hline
		No Consecutive Blocks (NCB)   & \makecell[l]{any clinician should not work \\
			two consecutive blocks, across all services}                                   
		& Hard          \\ \hline
		No Consecutive Weekends (NCW) & \makecell[l]{any clinician should not work two
			consecutive \\ weekends}                                                       
		& Hard          \\ \hline
		Equal Weekends (EW)          & \makecell[l]{weekends should be equally
			distributed \\ between clinicians}                                              
		& Hard          \\ \hline
		Equal Holidays (EH)          & \makecell[l]{long weekends should be equally
			distributed \\ between clinicians}                                              
		& Hard          \\ \hline
		Block Requests (BR)         & \makecell[l]{each clinician can request to be
			off service \\ during certain blocks throughout the year}                       
		& Soft          \\ \hline
		Weekend Requests (WR)        & \makecell[l]{each clinician can request to be
			off service \\ during certain weekends throughout the year}                     
		& Soft          \\ \hline
		Block-Weekend Adjacency (BWA)  & \makecell[l]{the block and weekend
			assignments of a given \\ clinician should be adjacent}                         
		& Soft
	\end{tabular}
	\caption{Summary of the constraints for the clinician scheduling problem}\label{tbl:constraint-summary}
\end{table}

>>>>

At St.\ Michael's Hospital, the division of infectious diseases (ID) offers
general ID and HIV consultation on inpatients as two parallel services. Each
service provides clinical care throughout the year, during regular work weeks as
well as weekends and holidays. The clinicians in the division typically receive
a schedule in advance, outlining their on-call service dates for the full year.
In the yearly schedule each clinician is assigned to blocks of regular work
weeks and weekends. Each block corresponds to two consecutive work weeks. Apart
from long (holiday) weekends, a work week starts on Monday at 8 A.M. and ends on
Friday at 5 P.M. Accordingly, weekend service starts on Friday at 5 P.M., and
ends on Monday at 8 A.M. During the weekend, ID and HIV consultation services
are combined and provided by one clinician. During the regular work week the ID
and HIV services are led by one clinician each.

Several constraints are placed on the on-call assignments. First, each clinician
has limits on the number of blocks they can and must work during the year,
depending on the type of consultation. For instance, a clinician might have to
provide 3-5 blocks of general ID consultation as well as 2-3 blocks of HIV
consultation throughout the year. These limits may change from year to year as
the number of clinicians in the department changes. Moreover, the schedule
should not assign a clinician to work for two consecutive blocks or two
consecutive weekends in a row. The schedule should also distribute both regular
and holiday weekends equally among all clinicians.

@@

\begin{table}[h]
	\centering
	\caption{Summary of the constraints for the clinician scheduling problem}%
  \label{tbl:constraint-summary}
  % JK: table caption should go at the top.
  %     The spacing can be funny so I use:
  %     \usepackage{caption}\captionsetup[table]{skip=1em}
  %     also labels should be on their own line.
  % JK: below: I would consider making the 2-3 letter abbreviations its own column
	\begin{tabular}{ l l l }
		\toprule % JK: different line weights support table structure
		\textbf{Constraint Name} & \textbf{Description}                               
		&
		\textbf{Type} \\ \midrule
		Block Coverage (BC)           & \makecell[l]{each service needs to have
			exactly \\ one clinician that covers any given block}                           
		& Hard          \\ \hline
		Weekend Coverage (WC)        & \makecell[l]{every weekend needs to have
			exactly \\ one clinician that covers it}                                        
		& Hard          \\ \hline
		Min/Max (MM)                  & \makecell[l]{for a given service, each
			clinician can only \\ work between the minimum and maximum \\ number of allowed
			blocks} & Hard          \\ \hline
		No Consecutive Blocks (NCB)   & \makecell[l]{any clinician should not work \\
			two consecutive blocks, across all services}                                   
		& Hard          \\ \hline
		No Consecutive Weekends (NCW) & \makecell[l]{any clinician should not work two
			consecutive \\ weekends}                                                       
		& Hard          \\ \hline
		Equal Weekends (EW)          & \makecell[l]{weekends should be equally
			distributed \\ between clinicians}                                              
		& Hard          \\ \hline
		Equal Holidays (EH)          & \makecell[l]{long weekends should be equally
			distributed \\ between clinicians}                                              
		& Hard          \\ \hline
		Block Requests (BR)         & \makecell[l]{each clinician can request to be
			off service \\ during certain blocks throughout the year}                       
		& Soft          \\ \hline
		Weekend Requests (WR)        & \makecell[l]{each clinician can request to be
			off service \\ during certain weekends throughout the year}                     
		& Soft          \\ \hline
		Block-Weekend Adjacency (BWA)  & \makecell[l]{the block and weekend
			assignments of a given \\ clinician should be adjacent}                         
		& Soft
    \\\bottomrule % JK: I think usually table bottom also has a rule
	\end{tabular}
  % JK: please define Hard & Soft constraints in a footnote
\end{table}

### 2019-09-13 paper/paper.results.tex
paragraph spacing
tables & figures cleanup

\subsection{Infectious Diseases Division}  %main comment  = I think use past
%tense in writing
We used clinician time-off requests and minimum/maximum requirements from
2015-2018 as input data for the ILP problem. Table~\ref{tbl:2018-schedule-comparison} presents the optimal schedule generated using
the software as well as the manually-created schedule for data from 2018. The
schedule is color-coded to distinguish between the different clinicians
assigned. \\

\include{tbl/2018_schedule_comparison}

First, we evaluated the software-generated schedule by comparing it with the
manually-created schedule. Specifically, we examined the adherence of each
schedule to the constraints presented in Table~\ref{tbl:constraint-summary}. As
shown in Table~\ref{tbl:constraints-comparison}, the optimal schedule generated
using the software satisfied all mandatory constraints. In contrast, the
manually-created schedule was not able to satisfy all mandatory constraints. In
particular, we see that the manual schedule assigned clinicians to multiple
consecutive blocks in all four years. Moreover, the manual schedule did not have
an equal distribution of weekends and holidays for all four years of data.
Evaluating the objectives, we see that LP outperforms the manually created
schedule in all four years, by accommodating almost all time-off requests and
ensuring that weekends are always assigned close to blocks. \\

\include{tbl/constraints_comparison}

\subsection{Simulations}
We then examined the properties of the ILP problem using simulated data.
Specifically, we examined the influence of expanding the following parameters on
the runtime of the algorithm: number of clinicians; number of services offered
in the division; amount of time-off requests per clinician throughout the year.
We also examined the effect of a longer time-horizon on the runtime. \\

The effect of an increasing number of requests per clinician on the runtime of
the ILP solver is shown in Figure~\ref{fig:runtime-requests}. We executed the
program on a department with 10 clinicians offering two services, similar to the
department at St.\ Michael's Hospital. For simplicity, each clinician was
configured with $\abs{\mathcal{BR}} = \{0, 5, 10, 15\}$ non-overlapping block
requests. The increase in the number of requests did not affect the runtime of
the ILP solver, indicating that it can accommodate a lot of flexibility in
clinician requests. Moreover, we see that all four runs were completed in under
1 second. \\

\begin{figure}[h]
	\centering
	\includegraphics[scale=.5]{fig/runtime_requests}
	\caption{Runtime of ILP solver with an increasing number of requests per
		clinician}\label{fig:runtime-requests}
\end{figure}

Figure~\ref{fig:runtime-blocks} presents the change in runtime when increasing
the number of 2-week blocks $B = \{26, 52, 78, 104\}$ in a department with 10
clinicians offering two services. There does not seem to be a strong effect on
runtime when we lengthen the time-horizon of the schedule, and in fact the
solver was able to find an optimal schedule for all time horizons within 3
seconds. \\ % 'very reasonable amount of time" is vague. what is reasonable?
%better to give a range in value.

\begin{figure}[h]
	\centering
	\includegraphics[scale=.5]{fig/runtime_blocks}
	\caption{Runtime of ILP solver with an increasing number of 2-week blocks}\label{fig:runtime-blocks}
\end{figure}

The effect of increasing the number of services %divisions %do you mean
%services?
on the runtime of the program is shown in Table~\ref{tbl:runtime-services-clinicians-comparison}. We executed the algorithm for
$S = \{1, 2, 3\}$ total services and $C = \{10, 20, 30, 50\}$ clinicians in
total across all services. For 2 concurrent services, a roster of 30 or more
clinicians becomes impractical to schedule, as searching for a solution required
over 24 hours. We saw similar issues for a roster of 20 or more clinicians
assigned to a division with 3 concurrent services. However, when relaxing the
NCB constraint, we saw a great improvement in runtime for divisions with 2 and 3
services, and we were able to generate a schedule with upwards of 50 clinicians
in under 1.5 seconds. \\  %what does division mean? do you mean services?

\include{tbl/runtime_services_clinicians}

>>>>

\subsection{Infectious Diseases Division}  %main comment  = I think use past
%tense in writing
We used clinician time-off requests and minimum/maximum requirements from
2015-2018 as input data for the ILP problem. Table~\ref{tbl:2018-schedule-comparison} presents the optimal schedule generated using
the software as well as the manually-created schedule for data from 2018. The
schedule is color-coded to distinguish between the different clinicians
assigned.

\input{tbl/2018_schedule_comparison}
% JK: I think use input not include here to avoid unnecessary page breaks
%     https://tex.stackexchange.com/questions/246

First, we evaluated the software-generated schedule by comparing it with the
manually-created schedule. Specifically, we examined the adherence of each
schedule to the constraints presented in Table~\ref{tbl:constraint-summary}. As
shown in Table~\ref{tbl:constraints-comparison}, the optimal schedule generated
using the software satisfied all mandatory constraints. In contrast, the
manually-created schedule was not able to satisfy all mandatory constraints. In
particular, we see that the manual schedule assigned clinicians to multiple
consecutive blocks in all four years. Moreover, the manual schedule did not have
an equal distribution of weekends and holidays for all four years of data.
Evaluating the objectives, we see that LP outperforms the manually created
schedule in all four years, by accommodating almost all time-off requests and
ensuring that weekends are always assigned close to blocks.

\input{tbl/constraints_comparison}

\subsection{Simulations}
We then examined the properties of the ILP problem using simulated data.
Specifically, we examined the influence of expanding the following parameters on
the runtime of the algorithm: number of clinicians; number of services offered
in the division; amount of time-off requests per clinician throughout the year.
We also examined the effect of a longer time-horizon on the runtime.

The effect of an increasing number of requests per clinician on the runtime of
the ILP solver is shown in Figure~\ref{fig:runtime-requests}. We executed the
program on a department with 10 clinicians offering two services, similar to the
department at St.\ Michael's Hospital. For simplicity, each clinician was
configured with $\abs{\mathcal{BR}} = \{0, 5, 10, 15\}$ non-overlapping block
requests. The increase in the number of requests did not affect the runtime of
the ILP solver, indicating that it can accommodate a lot of flexibility in
clinician requests. Moreover, we see that all four runs were completed in under
1 second.

\begin{figure}[h]
	\centering
	\includegraphics[scale=.5]{fig/runtime_requests} % JK: can you export this to PDF or SVG to avoid pixelation?
	\caption{Runtime of ILP solver with an increasing number of requests per clinician}
  \label{fig:runtime-requests}
  % JK: The trend here is a bit confusing, since I think you're trying to say
  %     that the runtime is not affected by the number of constraints,
  %     but there is noise in these results.
  %     I would re-run this 10-100 times and average the results so that (I expect)
  %     the mean trend is actually flat. You could (should) then also present the runtimes
  %     as box-plots to show that the trend is less than the variability.
\end{figure}

Figure~\ref{fig:runtime-blocks} presents the change in runtime when increasing
the number of 2-week blocks $B = \{26, 52, 78, 104\}$ in a department with 10
clinicians offering two services. There does not seem to be a strong effect on
runtime when we lengthen the time-horizon of the schedule, and in fact the
solver was able to find an optimal schedule for all time horizons within 3
seconds. % 'very reasonable amount of time" is vague. what is reasonable?
%better to give a range in value.

\begin{figure}[h]
	\centering
	\includegraphics[scale=.5]{fig/runtime_blocks}
	\caption{Runtime of ILP solver with an increasing number of 2-week blocks}%
  \label{fig:runtime-blocks}
  % JK: same comment as above re. trends -- are you trying to show one?
  % JK: Thess figures could also possible be Figure 1. a) and b) 
\end{figure}

The effect of increasing the number of services %divisions %do you mean
%services?
on the runtime of the program is shown in Table~\ref{tbl:runtime-services-clinicians-comparison}. We executed the algorithm for
$S = \{1, 2, 3\}$ total services and $C = \{10, 20, 30, 50\}$ clinicians in
total across all services. For 2 concurrent services, a roster of 30 or more
clinicians becomes impractical to schedule, as searching for a solution required
over 24 hours. We saw similar issues for a roster of 20 or more clinicians
assigned to a division with 3 concurrent services. However, when relaxing the
NCB constraint, we saw a great improvement in runtime for divisions with 2 and 3
services, and we were able to generate a schedule with upwards of 50 clinicians
in under 1.5 seconds.  %what does division mean? do you mean services?

\input{tbl/runtime_services_clinicians}
